---
params:
  title: ''
  subtitle: ''
  authors: ''
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r getRunDateTime()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    fig_width: 5
always_allow_html: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}

# Load Packages ----
rmdLibs <- c("openair", "skimr")

dkUtils::loadLibraries(rmdLibs)

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)

# Log compile time:
myParams$startTime <- proc.time()

# Parameters ----
# set xlim for plotly to reduce plot size & load speed
myParams$xlimMinDateTime <- lubridate::as_datetime("2018-01-01 00:00:00")
myParams$xlimMaxDateTime <- lubridate::as_datetime("2020-06-01 00:00:00")
myParams$xlimMinDate <- lubridate::as_date("2018-01-01")
myParams$xlimMaxDate <- lubridate::as_date("2020-06-01")

myParams$oneYearAgo <- lubridate::as_datetime(now() - (365*24*60*60))

# set values for annotations
myParams$lockDownStartDate <- as.Date("2020-03-24")
myParams$lockDownStartDateTime <- lubridate::as_datetime("2020-03-24 00:00:00")
myParams$lockDownEndDate <- lubridate::today()
myParams$lockDownEndDateTime <- lubridate::now()

myParams$recentCutDate <- lubridate::as_date("2020-03-01")

myParams$gamCap <- "Trend line = Generalized additive model (gam) with integrated smoothness estimation"
myParams$lockdownCap <- "\nColoured rectangle = UK covid lockdown to date"
myParams$weekendCap <- "\nShaded rectangle = weekends"

myParams$myAlpha <- 0.1
myParams$vLineAlpha <- 0.4
myParams$vLineCol <- "red" # http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
myParams$myTextSize <- 4


```

# About

## Code

 * https://github.com/CfSOtago/airQual/commits/master

## Citation

If you wish to refer to any of the material from this report please cite as:

 * Anderson, B., (`r format(Sys.time(), format = "%Y")`) `r params$title`: `r params$subtitle` , [Sustainable Energy Research Group](http://www.energy.soton.ac.uk), University of Southampton: Southampton, UK.

Report circulation:

 * Public
 
This work is (c) `r format(Sys.time(), format = "%Y")` the University of Southampton.

## Disclaimer

I usually do energy demand research but in the absence of access to real time energy demand data on lockdown (unlike during the [World Cup](http://www.energy.soton.ac.uk/the-end-of-the-world-cup/)), I'm looking at other things.
 
# Introduction

We use the excellent openair package [@openair] to download the AURN data and create the wind and pollution roses for each of the following Southampton sites:

 * [SA33](https://uk-air.defra.gov.uk/networks/site-info?site_id=SA33&view=View) - A33 near docks
 * [SOUT](https://uk-air.defra.gov.uk/networks/site-info?site_id=SOUT&view=View) - City Centre

Year and site is given in the legend label.

This analysis uses data for Southampton downloaded from:

 * [AURN](https://uk-air.defra.gov.uk/networks/network-info?view=aurn)
 
> The data supplied to AURN by UK monitoring sites such as the ones in Southampton is  [ratified](https://uk-air.defra.gov.uk/assets/documents/Data_Validation_and_Ratification_Process_Apr_2017.pdf) to check for outliers and instrument/measurement error. However, AURN data less than six months old has not undergone this process. Be warned.

# Data

Downloaded from AURN.


```{r prepData}

sites <- c("SA33", "SOUT")

# use:
# windRose(mydata, type = "year", layout = c(4, 2))
# https://www.rdocumentation.org/packages/openair/versions/2.7-0


  # wide form
    df_SA33 <- openair::importAURN(
      site = "SA33",
      year = 2000:2020,
      pollutant = "all",
      hc = FALSE,
      to_narrow = FALSE, # produces wide form data
      verbose = TRUE
    )
  
    df_SOUT <- openair::importAURN(
      site = "SOUT",
      year = 2000:2020,
      pollutant = "all",
      hc = FALSE,
      to_narrow = FALSE, # produces wide form data
      verbose = TRUE
    )
```


# Site: SA33

A33 - near docks: [map](https://uk-air.defra.gov.uk/networks/site-info?site_id=SA33&view=View)

```{r descSA33}

skimr::skim(df_SA33)
```

## Wind roses

```{r windRoseSA33}
# make a compass rose for this site for this year
openair::windRose(df_SA33, type = "year", 
                              key.header = paste0("Southampton site: ", s),
                              layout = c(3,2))
```

## Pollution roses

```{r pollutionRoseSA33}
# make a compass rose for this site for this year
openair::pollutionRose(df_SA33, type = "year", 
                              key.header = paste0("Southampton site: ", s),
                              layout = c(3,2))
```

# Site: SOUT

City centre: [map](https://uk-air.defra.gov.uk/networks/site-info?site_id=SOUT&view=View)

```{r descSOUT}

skimr::skim(df_SOUT)
```

## Wind roses

```{r windRoseSOUT}
# make a compass rose for this site for this year
openair::windRose(df_SOUT, type = "year", 
                              key.header = paste0("Southampton site: ", s),
                              layout = c(3,2))
```

## Pollution roses

```{r pollutionRoseSOUT}
# make a compass rose for this site for this year
openair::pollutionRose(df_SOUT, type = "year", 
                              key.header = paste0("Southampton site: ", s),
                              layout = c(3,2))
```


# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - myParams$startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used:

 * here - [@here]
 * openair - [@openair]
            
# References

