---
params:
  title: ''
  subtitle: ''
  authors: ''
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r getRunDateTime()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    fig_width: 5
always_allow_html: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}

# Load Packages ----
rmdLibs <- c("openair")

dkUtils::loadLibraries(rmdLibs)

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)

# Log compile time:
myParams$startTime <- proc.time()

# Parameters ----
# set xlim for plotly to reduce plot size & load speed
myParams$xlimMinDateTime <- lubridate::as_datetime("2018-01-01 00:00:00")
myParams$xlimMaxDateTime <- lubridate::as_datetime("2020-06-01 00:00:00")
myParams$xlimMinDate <- lubridate::as_date("2018-01-01")
myParams$xlimMaxDate <- lubridate::as_date("2020-06-01")

myParams$oneYearAgo <- lubridate::as_datetime(now() - (365*24*60*60))

# set values for annotations
myParams$lockDownStartDate <- as.Date("2020-03-24")
myParams$lockDownStartDateTime <- lubridate::as_datetime("2020-03-24 00:00:00")
myParams$lockDownEndDate <- lubridate::today()
myParams$lockDownEndDateTime <- lubridate::now()

myParams$recentCutDate <- lubridate::as_date("2020-03-01")

myParams$gamCap <- "Trend line = Generalized additive model (gam) with integrated smoothness estimation"
myParams$lockdownCap <- "\nColoured rectangle = UK covid lockdown to date"
myParams$weekendCap <- "\nShaded rectangle = weekends"

myParams$myAlpha <- 0.1
myParams$vLineAlpha <- 0.4
myParams$vLineCol <- "red" # http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
myParams$myTextSize <- 4


```

# About

## Code

 * https://github.com/CfSOtago/airQual/commits/master

## Citation

If you wish to refer to any of the material from this report please cite as:

 * Anderson, B., (`r format(Sys.time(), format = "%Y")`) `r params$title`: `r params$subtitle` , [Sustainable Energy Research Group](http://www.energy.soton.ac.uk), University of Southampton: Southampton, UK.

Report circulation:

 * Public
 
This work is (c) `r format(Sys.time(), format = "%Y")` the University of Southampton.

## Disclaimer

> I usually do energy demand research but in the absence of access to real time demand data on lockdown (unlike during the [World Cup](http://www.energy.soton.ac.uk/the-end-of-the-world-cup/)) I'm looking at other things.
 
# Introduction

Data for Southampton downloaded from :

 * [AURN](https://uk-air.defra.gov.uk/networks/network-info?view=aurn). 
 
The data that goes to AURN is then [ratifified](https://uk-air.defra.gov.uk/assets/documents/Data_Validation_and_Ratification_Process_Apr_2017.pdf) to check for outliers and instrument/measurement error. AURN data less than six months old has not undergone this process.

# Wind roses

We use the excellent openair package [@openair] to download the AURN data and create the wind and pollution roses for each site.

```{r prepData}

years <- c("2016", "2017", "2018","2019","2020")
sites <- c("SA33", "SOUT")

for(y in years){
  
  # wide form
  for(s in sites){
    df <- openair::importAURN(
      site = s,
      year = y,
      pollutant = "all",
      hc = FALSE,
      to_narrow = FALSE, # produces wide form data
      verbose = TRUE
    )

    # make a compass rose for this site for this year
    message("Wind rose: ", y, " (Southampton - ", s, ")")
    lab <- paste0("Southampton: ", s, " (",y,")")
    openair::windRose(df , key.header = paste0("Southampton site: ", s, " (",y,")"))
    message("Pollution rose: ", y, " (Southampton - ", s, ")")
    openair::pollutionRose(df, key.header = paste0("Southampton site: ", s, " (",y,")"))
  }

}

# done ----

```


# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - myParams$startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used:

 * here - [@here]
 * openair - [@openair]
            
# References

