---
params:
  title: ''
  subtitle: ''
  authors: ''
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r getRunDateTime()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    fig_width: 5
always_allow_html: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}

# Load Packages ----
rmdLibs <- c("ggplot2",
            "kableExtra",
            "plotly",
            "skimr",
            "viridis")

dkUtils::loadLibraries(rmdLibs)

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)

# Log compile time:
myParams$startTime <- proc.time()

# Parameters ----
# set xlim for plotly to reduce plot size & load speed
myParams$xlimMinDateTime <- lubridate::as_datetime("2018-01-01 00:00:00")
myParams$xlimMaxDateTime <- lubridate::as_datetime("2020-06-01 00:00:00")
myParams$xlimMinDate <- lubridate::as_date("2018-01-01")
myParams$xlimMaxDate <- lubridate::as_date("2020-06-01")

myParams$oneYearAgo <- lubridate::as_datetime(now() - (365*24*60*60))

# set values for annotations
myParams$lockDownStartDate <- as.Date("2020-03-24")
myParams$lockDownStartDateTime <- lubridate::as_datetime("2020-03-24 00:00:00")
myParams$lockDownEndDate <- lubridate::today()
myParams$lockDownEndDateTime <- lubridate::now()

myParams$recentCutDate <- lubridate::as_date("2020-03-01")

myParams$gamCap <- "Trend line = Generalized additive model (gam) with integrated smoothness estimation"
myParams$lockdownCap <- "\nColoured rectangle = UK covid lockdown to date"
myParams$weekendCap <- "\nShaded rectangle = weekends"

myParams$myAlpha <- 0.1
myParams$vLineAlpha <- 0.4
myParams$vLineCol <- "red" # http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
myParams$myTextSize <- 4

# Functions ----
makeDotPlot <- function(dt, xVar, yVar, byVar, yLab){
  p <- ggplot2::ggplot(dt, aes(x = get(xVar), 
                               y = get(yVar),
                               colour = get(byVar),
                               alpha = 1/100 # https://ggplot2.tidyverse.org/reference/geom_point.html
                               )
                       ) +
    geom_point(size = 1, # small
               shape = 4 # small cross shape
               ) +
    scale_colour_viridis_d(name = eval(byVar)) +
    guides(alpha = FALSE) + # remove alpha legend 
    labs(x = "Time",
         y = eval(yLab)) +  
    theme(legend.position="bottom") +
    guides(colour = guide_legend(nrow = 2)) # forces 2 rows to legend
  return(p)
}

makeTilePlot <- function(dt, xVar, yVar, fillVar, yLab){
  p <- ggplot2::ggplot(dt, aes(x = get(xVar), 
                               y = get(yVar),
                               fill = get(fillVar)
                               )
                       ) +
    geom_tile() +
    scale_fill_continuous(low = "green", high = "red", name = "Value") +
    labs(x = "Time",
         y = yLab) +  
    theme(legend.position="bottom")
  return(p)
}

addLockdownDate <- function(p){
  # assumes p has x = obsDate
  # p <- p + annotate("text", x = myParams$lockDownStartDate, 
  #            y = yMax * 0.4, angle = 10,size = myParams$myTextSize,
  #            label = "UK covid lockdown to date", hjust = 0.5)
  p <- p + annotate("rect", xmin = myParams$lockDownStartDate,
             xmax = myParams$lockDownEndDate, 
             ymin = yMin, ymax = yMax, 
             alpha = myParams$myAlpha, 
             fill = myParams$vLineCol, 
             colour = myParams$vLineCol)
  return(p)
}

addLockdownDateTime <- function(p){
  # assumes p has x = obsDateTime
  # p <- p + annotate("text", x = myParams$lockDownStartDateTime, 
  #            y = yMax * 0.4, angle = 10,size = myParams$myTextSize,
  #            label = "UK covid lockdown to date", hjust = 0.5)
  p <- p + annotate("rect", xmin = myParams$lockDownStartDateTime,
             xmax = myParams$lockDownEndDateTime, 
             ymin = yMin, ymax = yMax, 
             alpha = myParams$myAlpha, 
             fill = myParams$vLineCol, 
             colour = myParams$vLineCol) 
    
  return(p)
}

# only makes sense to use these for x axis covering March onwards
myParams$weAlpha <- 0.3
myParams$weFill <- "grey50"
addWeekendsDate <- function(p){
  p <- p + annotate("rect", xmin = as.Date("2020-03-07"),
                    xmax = as.Date("2020-03-08"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = as.Date("2020-03-14"),
                    xmax = as.Date("2020-03-15"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = as.Date("2020-03-21"),
                    xmax = as.Date("2020-03-22"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = as.Date("2020-03-28"),
                    xmax = as.Date("2020-03-29"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = as.Date("2020-04-04"),
                    xmax = as.Date("2020-04-05"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  return(p)
}

addWeekendsDateTime <- function(p){
   p <- p + annotate("rect", xmin = lubridate::as_datetime("2020-03-07 00:00:00"),
                    xmax = lubridate::as_datetime("2020-03-08 23:59:59"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = lubridate::as_datetime("2020-03-14 00:00:00"),
                    xmax = lubridate::as_datetime("2020-03-15 23:59:59"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = lubridate::as_datetime("2020-03-21 00:00:00"),
                    xmax = lubridate::as_datetime("2020-03-22 23:59:59"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = lubridate::as_datetime("2020-03-28 00:00:00"),
                    xmax = lubridate::as_datetime("2020-03-29 23:59:59"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  p <- p + annotate("rect", xmin = lubridate::as_datetime("2020-04-04 00:00:00"),
                    xmax = lubridate::as_datetime("2020-04-05 23:59:59"),
                    ymin = yMin, ymax = yMax,
                    alpha = myParams$weAlpha, fill = myParams$weFill)
  return(p)
}
```

# About

## Purpose

Extracting data for Southampton A33 location for input to modelling.

## Code

 * https://github.com/CfSOtago/airQual/commits/master
 
# Data

Data for Southampton downloaded from :

 * https://uk-air.defra.gov.uk/networks/network-info?view=aurn

Southampton City Council collects various forms of air quality data at the sites shown in \@ref(tab:showSites). Some of these sites feed data to AURN. 

The AURN data then undergoes a manual check and [ratification](https://uk-air.defra.gov.uk/assets/documents/Data_Validation_and_Ratification_Process_Apr_2017.pdf) process. Data that is less than 6 months old has _not_ undergone this process.

# Extract specification

Pollutants: CO, NO2, NOX, PM10, PM2.5 (ideally, or any subset – PM10 less useful as not such a strong road signal)

Dates: 

 * 11 Feb – 5 May 2019 - Training dataset
 * 11 Feb – 23 March 2020 – Model testing
 * 24 Mar – now	- Reduction prediction

Model runs on 10-minute averages or hourly averages – would be useful to have both.

Ultimately need to create a dataset with: Year, month, day, hour, minute (0,10,20,30,40,50), weekday/weekend, wind speed, wind direction, upwind/downwind, average pollutant concentration


> wind speed etc available on AURN data (not raw Southampton data)

For much more detailed analysis see a longer and very messy [data report](https://cfsotago.github.io/airQual/sccAirQualExplore_Exploring%20the%20SSC%20and%20AURN%20data.html).

```{r showSites}

aurnDT <- aurnDT[obsDate > as.Date("2018-01-01")] # for speed

aurnDT[, pollutant := ifelse( pollutant == "wd", "windDirection", pollutant)]
aurnDT[, pollutant := ifelse( pollutant == "ws", "windSpeed", pollutant)]

t <- table(aurnDT$pollutant,aurnDT$site)

kableExtra::kable(t, caption = "Dates where data != NA by site and measure",
                  digits = 2) %>%
  kable_styling()
```

Site locations:

 * Southampton A33: https://uk-air.defra.gov.uk/networks/site-info?site_id=SA33&view=View
 * Southampton Centre: https://uk-air.defra.gov.uk/networks/site-info?site_id=SOUT&view=View
 
# Nitrogen Dioxide (no2)


```{r No2Summary}
yLab <- "Nitrogen Dioxide (ug/m3)"
no2dt <- aurnDT[pollutant == "no2"]
```

Figure \@ref(fig:testo2) shows the availability of this data.

```{r testo2, fig.cap="Nitrogen Dioxide data availability and levels over time"}
# dt,xvar, yvar,fillVar, yLab
p <- makeTilePlot(no2dt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```

# Oxides of Nitrogen (nox)

```{r NoxSummary}
yLab <- "Oxides of Nitrogen (ug/m3)"
noxdt <- aurnDT[pollutant == "nox"]
```

Figure \@ref(fig:testNox) shows the availability of this data over time.

```{r testNox, fig.cap="Oxides of nitrogen data availability and levels over time"}

# dt,xvar, yvar,fillVar, yLab
p <- makeTilePlot(noxdt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```


# Sulphour Dioxide

```{r So2Summary}
yLab <- "Sulphour Dioxide (ug/m3)"
so2dt <- aurnDT[pollutant == "so2"]

```


Figure \@ref(fig:testS02) shows the availability of this data over time.

```{r testS02, fig.cap="Sulphour Dioxide data availability and levels over time"}

# dt,xvar, yvar,fillVar, yLab
p <- makeTilePlot(so2dt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```

# Ozone


```{r o3Summary}
yLab <- "Ozone (ug/m3)"
o3dt <- aurnDT[pollutant == "o3"]
```

Figure \@ref(fig:testo3) shows the most recent hourly data.

```{r testo3, fig.cap = "Availability and level of o3 data over time"}

p <- makeTilePlot(o3dt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```

# PM 10

```{r pm10Summary}
yLab <- "PM 10 (ug/m3)"
pm10dt <- aurnDT[pollutant == "pm10"]

```

Figure \@ref(fig:pm10test) shows the availability of data over time.

```{r pm10test, fig.cap = "Availability and level of o3 data over time"}

p <- makeTilePlot(pm10dt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```

# PM 2.5

```{r pm25Summary}
yLab <- "PM 2.5 (ug/m3)"
pm25dt <- aurnDT[pollutant == "pm2.5"]

```

Figure \@ref(fig:pm25test) shows the availability of data over time.

```{r pm25test, fig.cap = "Availability and level of o3 data over time"}

p <- makeTilePlot(pm25dt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```

## Wind speed

```{r WSSummary}
yLab <- "Wind speed (m/s)"
wsdt <- aurnDT[pollutant == "windSpeed"]

```

Figure \@ref(fig:wstest) shows the availability of data over time.

```{r wstest, fig.cap = "Availability and level of wind speed data over time"}

p <- makeTilePlot(wsdt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```

## Wind direction

```{r wdSummary}
yLab <- "Wind direction (deg)"
wddt <- aurnDT[pollutant == "windDirection"]

```

Figure \@ref(fig:wdtest) shows the availability of data over time.

```{r wdtest, fig.cap = "Availability and level of wind direction data over time"}

p <- makeTilePlot(wddt, xVar = "dateTimeUTC", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```

# Save data

Save long form data to [data](../data/) folder.

```{r saveData}
aurnDT[, weekDay := lubridate::wday(dateTimeUTC, label = TRUE, abbr = TRUE)]
f <- paste0(here::here(), "/data/sotonExtract2018_2020.csv")
data.table::fwrite(aurnDT, f)
```

Saved data description:

```{r dataDesc}
skimr::skim(aurnDT)
```

# Annex

# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - myParams$startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used:

 * data.table - [@data.table]
 * ggplot2 - [@ggplot2]
 * here - [@here]
 * kableExtra - [@kableExtra]
 * lubridate - [@lubridate]
 * skimr - [@skimr]
 * viridis - [@viridis]
            
# References

