---
params:
  myTitle: 'Air Quality in New Zealand'
  mySubTitle: 'Exploring "official" data'
title: "`r params$myTitle`"
subtitle: "`r params$mySubTitle`"
author: "Ben Anderson (b.anderson@soton.ac.uk `@dataknut`)"
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    collapsed: FALSE
    toc_float: yes
    highlight: tango
    code_folding: hide
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}

# Load Packages ----
library(dkUtils)

myLibs <- c("data.table",
            "ggplot2",
            "here",
            "kableExtra",
            "lubridate",
            "plotly",
            "skimr")

dkUtils::loadLibraries(myLibs)              # Load script specific packages

# Project Settings ----
projLoc <- here::here()

annualThreshold <- 10 # WHO
dailyThreshold <- 25 # WHO

dPath <- "~/Data/NZ_mfe/airQuality/"
pm2.5File <- "mfe-pm25-concentrations-200817-CSV/pm25-concentrations-200817.csv"
pm10File <- "mfe-pm10-concentrations-200617-CSV/pm10-concentrations-200617.csv"

pmURL <- "https://data.mfe.govt.nz/data/category/air/"
  

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)

# Log compile time:
startTime <- proc.time()

# Functions ----
makeTilePlot <- function(dt,yvar,byvar){
  p <- ggplot2::ggplot(dt, aes(x = ba_date, 
                               y = get(byvar), 
                               fill = get(yvar)
                               )
                       ) +
    geom_tile() +
    scale_fill_continuous(low = "green", high = "red") +
    labs(x = "Date")
  return(p)
}

makeLinePlot <- function(dt,yvar,byvar){
  p <- ggplot2::ggplot(dt, aes(x = ba_date, 
                               y = get(yvar),
                               colour = get(byvar)
                               )
                       ) +
    geom_line() +
    labs(x = "Date")
  return(p)
}
```

# About

## Contributions

Please note that authorship is alphabetical. Contributions are listed below - see [github](https://github.com/CfSOtago/airQual/commits/master) for details and who to blame for what :-).

Ben Anderson (b.anderson@soton.ac.uk `@dataknut`)

## Code

 * https://github.com/CfSOtago/airQual/commits/master

## Citation

If you wish to refer to any of the material from this report please cite as:

 * Anderson, B., (`r format(Sys.time(), format = "%Y")`) `r params$myTitle`:`r params$myTitle` , University of Southampton: Southampton, UK.

Report circulation:

 * Public.
 
Report purpose: 
 
 * to explore `official` New Zealand Air Quality data

This work has received funding from the European Union’s Horizon 2020 research and innovation programme under the [Marie Skłodowska-Curie](http://ec.europa.eu/research/mariecurieactions/index_en.htm) grant agreement No 700386 ([SPATIALEC](http://www.energy.soton.ac.uk/tag/spatialec/)).

This work is (c) `r format(Sys.time(), format = "%Y")` the University of Southampton.
 
# Introduction

# PM 10 data

PM 10 data: has more sensors and wider coverage.

Data source: `r pmURL`

Data file: `r pm10File`

```{r loadPM10}

df <- paste0(dPath, pm10File)

pm10dt <- data.table::fread(df)
pm10dt[, ba_date := lubridate::as_date(date)]
# the data is daily but there may be gaps?
pm10dt[, council.site := paste0(council, ".", site)]
```

Overall there are:

 * `r uniqueN(pm10dt$site)` sites spread over
 * `r uniqueN(pm10dt$council)` councils

```{r pm10TestData, fig.height = 8, fig.cap="Test data values by date and site"}
# looks like daily data with gaps
p <- makeTilePlot(pm10dt, yvar = "pm10", byvar = "council.site")
p + labs(y = "pm10") + 
  guides(fill=guide_legend(title="pm10"))
```

```{r pm10TestDataPlotly, fig.height = 8, fig.cap="Test data values by date and site"}
# looks like daily data with gaps
p <- makeLinePlot(pm10dt,yvar = "pm10",byvar = "council.site" )
p <- p + labs(y = "pm10") +
  guides(colour=guide_legend(title="pm10"))

plotly::ggplotly(p)
```


# PM 2.5 data

PM 2.5 data: has fewer sensors and less coverage.

Data source: `r pmURL`

Data file: `r pm2.5File`

```{r loadPM2.5}

df <- paste0(dPath, pm2.5File)

pm2.5dt <- data.table::fread(df)
pm2.5dt[, ba_date := lubridate::as_date(date)]
# the data is daily but there may be gaps?
pm2.5dt[, council.site := paste0(council, ".", site)]
```

Overall there are:

 * `r uniqueN(pm2.5dt$site)` sites spread over
 * `r uniqueN(pm2.5dt$council)` councils
 
 
```{r pm2.5TestData, fig.height = 8, fig.cap="Test data values by date and site"}
# looks like daily data with gaps
p <- makeTilePlot(pm2.5dt, yvar = "pm2_5", byvar = "council.site")
p + labs(y = "pm2_5") + 
  guides(fill=guide_legend(title="pm2_5"))
```

```{r pm2.5TestDataPlotly, fig.height = 8, fig.cap="Test data values by date and site"}
# looks like daily data with gaps
p <- makeLinePlot(pm2.5dt,yvar = "pm2_5",byvar = "council.site" )
p <- p + labs(y = "pm2_5") +
  guides(colour=guide_legend(title="pm2_5"))

plotly::ggplotly(p)
```

# Statistical Annex

## PM10

```{r skimPm10}
skimr::skim(pm10dt)
```


```{r pm10Sites}
# looks like daily data with gaps
t <- pm10dt[, .(nObs = .N), keyby = .(council, site)]

kableExtra::kable(t, caption = "N obs at sites (PM10)") %>%
  kable_styling()
  
```

## PM2.5

```{r skimPm2.5}
skimr::skim(pm2.5dt)
```


```{r pm2.5Sites}
# looks like daily data with gaps
t <- pm2.5dt[, .(nObs = .N), keyby = .(council, site)]

kableExtra::kable(t, caption = "N obs at sites (PM2.5)") %>%
  kable_styling()
  
```

# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used:

 * data.table - [@data.table]
 * ggplot2 - [@ggplot2]
 * here - [@here]
 * kableExtra - [@kableExtra]
 * lubridate - [@lubridate]
 * plotly - [@plotly]
 * skimr - [@skimr]
            
# References

