---
params:
  title: "Air Quality in Southampton (UK)"
  subtitle: "Exploring the data (using AURN & openair)"
  authors: "Ben Anderson (b.anderson@soton.ac.uk `@dataknut`)"
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
  bookdown::pdf_document2:
    fig_caption: yes
    number_sections: yes
  bookdown::word_document2:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    fig_width: 5
always_allow_html: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}



# Load Packages ----
rmdLibs <- c("data.table",
             "ggplot2",
            "kableExtra",
            "plotly",
            "skimr")

dkUtils::loadLibraries(rmdLibs)

# load last in case of clashes
require(openair)

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)

# Log compile time:
startTime <- proc.time()

# Parameters ----
# set xlim for plotly to reduce plot size & load speed
xlimMinDateTime <- lubridate::as_datetime("2019-01-01 00:00:00")
xlimMaxDateTime <- lubridate::as_datetime("2020-01-01 00:00:00")
xlimMinDate <- lubridate::as_date("2019-01-01")
xlimMaxDate <- lubridate::as_date("2020-01-01")

# https://www.who.int/news-room/fact-sheets/detail/ambient-(outdoor)-air-quality-and-health
annualPm10Threshold_WHO <- 20
dailyPm10Threshold_WHO <- 50 
annualPm2.5Threshold_WHO <- 10
dailyPm2.5Threshold_WHO <- 25 
annualno2Threshold_WHO <- 40
hourlyno2Threshold_WHO <- 200 

makePlotly <- "_plotly" # '_plotly' -> yes - for plotly versions of charts

# set doPlotly
if(makePlotly == "_plotly"){
  doPlotly <- 1 # for easier if statements
} else {
  doPlotly <- 0
}

# Functions ----
makeDotPlot <- function(dt, xVar, yVar, byVar, yLab){
  p <- ggplot2::ggplot(dt, aes(x = get(xVar), 
                               y = get(yVar),
                               colour = get(byVar),
                               alpha = 1/100 # https://ggplot2.tidyverse.org/reference/geom_point.html
                               )
                       ) +
    geom_point(size = 1, # small
               shape = 4 # small cross shape
               ) +
    scale_color_discrete(name = eval(byVar)) +
    guides(alpha = FALSE) + # remove alpha legend 
    labs(x = "Time",
         y = eval(yLab)) +  
    theme(legend.position="bottom") +
    guides(colour = guide_legend(nrow = 2)) # forces 2 rows to legend
  return(p)
}

makeTilePlot <- function(dt, xVar, yVar, fillVar, yLab){
  p <- ggplot2::ggplot(dt, aes(x = get(xVar), 
                               y = get(yVar),
                               fill = get(fillVar)
                               )
                       ) +
    geom_tile() +
    scale_fill_continuous(low = "green", high = "red", name = "Value") +
    labs(x = "Time",
         y = yLab) +  
    theme(legend.position="bottom")
  return(p)
}

```

# About

## Contributions

Please note that authorship is alphabetical. Contributions are listed below - see [github](https://github.com/CfSOtago/airQual/commits/master) for details and who to blame for what :-).

 * Ben Anderson (b.anderson@soton.ac.uk `@dataknut`)

## Code

 * https://github.com/CfSOtago/airQual/commits/master

## Citation

If you wish to refer to any of the material from this report please cite as:

 * Anderson, B., (`r format(Sys.time(), format = "%Y")`) `r params$title`: `r params$subtitle` , University of Southampton: Southampton, UK.

Report circulation:

 * Public
 
Report purpose: 
 
 * to explore (official) Southampton City Council Air Quality data (http://southampton.my-air.uk) sourced from
 * to test the [openair](http://davidcarslaw.github.io/openair/) [@openair] package

This work is (c) `r format(Sys.time(), format = "%Y")` the University of Southampton.
 
# Introduction

```{r loadData}
dfW <- openair::importAURN(
  site = "SA33",
  year = 2019,
  pollutant = "all",
  hc = FALSE,
  meta = TRUE,
  to_narrow = FALSE, # produces long form data yay!
  verbose = TRUE # for now
)

# fails. it worked before
# dfL <- openair::importAURN(
#   site = "SA33",
#   year = 2019,
#   pollutant = "all",
#   hc = FALSE,
#   meta = TRUE,
#   to_narrow = TRUE, # produces long form data yay!
#   verbose = TRUE
# )

dtW <- data.table::as.data.table(dfW) # we like data.tables

```

Data downloaded from http://uk-air.defra.gov.uk/openair/R_data/ using `Åpenair::importAURN()`.

Southampton City Council collects various forms of air quality data at the sites shown in \@ref(tab:showSites). WHO publishes [information](https://www.who.int/news-room/fact-sheets/detail/ambient-(outdoor)-air-quality-and-health) on the health consequences and "acceptable" exposure levels for each of these.


```{r showSites}

lDT <- data.table::melt(dtW,
                     id.vars=c("site","date","code","latitude", "longitude", "site_type"),
                     measure.vars = c("no","no2","nox","pm10","nv10","v10",
                                      "ws","wd"),
                     value.name = "value" # varies 
  )

# remove NA
lDT <- lDT[!is.na(value)]

t <- lDT[,.(from = min(date),
           to = max(date),
           nObs = .N), keyby = .(site, variable)]

kableExtra::kable(t, caption = "Dates data available by site and measure",
                  digits = 2) %>%
  kable_styling()
```

# Summarise data

Summarise previously downloaded and processed data... Note that this may not be completely up to date.

```{r summariseData}

skimr::skim(dfW)

```

Table \@ref(tab:summariseData) gives an indication of the availability of the different measures.


# Analysis

In this section we present graphical analysis of the previoulsy downloaded data. Note this is just a snapshot of the data available.

## Nitrogen Dioxide


```{r no2Summary}
yLab <- "Nitrogen Dioxide (ug/m3)"

t <- lDT[variable == "no2", .(mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            min = min(value, na.rm = TRUE),
            max = max(value, na.rm = TRUE)), keyby = .(site)]
kableExtra::kable(t, caption = "Summary of no2 data") %>%
  kable_styling()
```

Table \@ref(tab:no2Summary) suggests that there may be a few (`r nrow(lDT[variable == "no2" & value < 0])`) negative values. These are summarised in \@ref(tab:testNegNo2) while Figure \@ref(fig:testNegNo2) shows the availability and levels of the pollutant data over time.

```{r testNegNo2, fig.cap="Nitrogen Dioxide data availability and levels over time"}

t <- head(lDT[variable == "no2" & value < 0], 10)
kableExtra::kable(t, caption = "Negative no2 values (up to first 6)") %>%
  kable_styling()

t <- table(lDT[variable == "no2" & value < 0, .(site)])
kableExtra::kable(t, caption = "Negative no2 values (count by site)") %>%
  kable_styling()

# dt,xvar, yvar,fillVar, yLab
p <- makeTilePlot(lDT[variable == "no2"], xVar = "date", yVar = "site",
                  fillVar = "value",
                  yLab = yLab)

p
```




```{r plotNo2Hourly, fig.cap="Nitrogen Dioxide levels, Southampton (hourly)"}
# p <- ggplot2::ggplot(dt, aes(x = obsDateTime, 
#                              y = nox2,
#                              colour = site,
#                              alpha = 0.1)) +
#   geom_point(shape=4, size = 1)

t <- lDT[variable == "no2" & value > 200][order(-value)]

kableExtra::kable(caption = paste0("Values greater than WHO threshold (NO2 > ", 
                                   hourlyno2Threshold_WHO , ")"), 
                  head(t, 10)) %>%
  kable_styling()

p <- makeDotPlot(lDT[variable == "no2"], 
                 xVar = "date", 
                 yVar = "value", 
                 byVar = "site", 
                 yLab = yLab)

p <- p + geom_hline(yintercept = hourlyno2Threshold_WHO) +
  labs(caption = "Reference line = WHO hourly guideline threshold")


if(doPlotly){
  p
  plotly::ggplotly(p + xlim(xlimMinDateTime, xlimMaxDateTime)) # interactive, xlimited 
} else {
  p
}

```

Figure \@ref(fig:plotNo2Hourly) shows hourly values for all sites. In the study period there were `r nrow(t)` hours when the hourly Nitrogen Dioxide level breached WHO guidelines. The worst 10 cases are shown in Table \@ref(tab:plotNo2Hourly).

```{r plotNo2Daily, fig.cap="Nitrogen Dioxide levels, Southampton (daily mean"}
lDT[, obsDate := lubridate::date(date)]

plotDT <- lDT[variable == "no2", .(mean = mean(value, na.rm = TRUE)),
             keyby = .(obsDate, site)]

p <- makeDotPlot(plotDT, 
                 xVar = "obsDate", 
                 yVar = "mean", 
                 byVar = "site", 
                 yLab = yLab)

p <- p +
  geom_smooth() + # add smoothed line
  labs(caption = "Trend line = Generalized additive model (gam) with integrated smoothness estimation")

if(doPlotly){
  p
  plotly::ggplotly(p + xlim(xlimMinDate, xlimMaxDate)) # interactive, xlimited # interactive
} else {
  p
}

```


Figure \@ref(fig:plotNo2Daily) shows daily mean values for all sites over time and includes smoother trend lines for each site.

Clearly the mean daily values show less variance (and less extremes) than the hourly data and there has also been a decreasing trend over time.

## openair tests

Wind rose

```{r windRose, fig.cap="Wind rose for Southampton A33, 2019"}
openair::windRose(dfW)
```


Pollution rose

```{r pollutionRose, fig.cap="Pollution rose for Southampton A33, 2019, hourly data"}
openair::pollutionRose(dfW, pollutant = "no2")
```

We get a slightly higher % of high measures when the wind is from the SE?

# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used:

 * data.table - [@data.table]
 * ggplot2 - [@ggplot2]
 * here - [@here]
 * kableExtra - [@kableExtra]
 * lubridate - [@lubridate]
 * openair - [@openair]
 * plotly - [@plotly]
 * skimr - [@skimr]
            
# References

